### 2017-06-01 03-sequential-design.tex

The MCMC-aMMSB algorithm that is the subject of this paper was initially
implemented in Python~\cite{LiAW15}, as is common
with machine learning researchers and practitioners.
We converted this implementation into an efficient sequential C++ application
to serve as a baseline for the parallel and distributed implementations.
This is certainly not one of our contributions to the parallelization
of the algorithm, but for a fair analysis of the parallel implementations
a good baseline is necessary.

The Python implementation
relied on numpy~\cite{numpy} to perform numerical computations efficiently.
Besides, the algorithm
made heavy use of Python data structures such as sets, dictionaries and
lists, which have no efficient numpy equivalent.

First, the Python code base was ported into equivalent C++ code.
The program structure remained unchanged and the Python data structures were
substituted with their direct C++ alternatives.
%
The next step was to remove a number of inefficiencies.
For example, one recurring idiom in the Python implementation
was an expression of the form $a^y b^{1-y}$, where $y$ is either 0 or~1.
% Therefore, the expression is simply a choice
% between $a$ for $y=1$ and $b$ for $y=0$. However, as the compiler cannot infer
% the value of $y$, both expressions $a$ and $b$ were computed and their
% resulting values were raised to the power $y$ and $1-y$ respectively.
We transformed such expressions into conditional expressions which
compute either $a$ or $b$, and avoid floating-point exponentiation.
%
% Similarly, we transformed some expressions to a more efficient form.
% For example, we replaced expressions of the form $a^{0.5}\times{b}^{0.5}$
% with $sqrt(ab)$.
Other optimizations were loop strength
reduction and common subexpression lifting.
%
% In summary, a multitude of localized optimizations were applied to achieve
% an efficient sequential baseline version, which is 1000..1500 times faster
% than the original Python code, see the next Section.
All transformations were rigorously tested to ensure they do
not modify the behavior of the algorithm.
%
\begin{comment}
Further, we replaced calls to the system's random
functions with a custom implementation of the random generator
\textit{xorshift\_128}~\cite{Marsaglia:2003:XR}. This way, random calls no longer
involve system calls, so we can support easy and fast multi-threaded random
calls by providing each thread with its private, differently seeded, random
generator.
\end{comment}

>>>>

The MCMC-aMMSB algorithm that is the subject of this paper was initially
implemented in Python~\cite{LiAW15}, as is common
in the machine learning community.
We converted this implementation into an efficient sequential C++ application
to serve as a baseline for the parallel and distributed implementations.
We certainly do not consider this to be a contribution to the parallelization
of the algorithm, but for a fair analysis of the parallel implementations
a good baseline is necessary.

The Python implementation
relied on Numpy~\cite{numpy} to perform numerical computations efficiently.
Besides, the algorithm
made heavy use of Python data structures such as sets, dictionaries and
lists, which have no efficient Numpy equivalent.

First, the Python code base was ported into equivalent C++ code.
The program structure remained unchanged and the Python data structures were
substituted with their direct C++ alternatives.
%
The next step was to remove a number of inefficiencies.
For example, one recurring idiom in the Python implementation
was a choice expression of the form $a^y b^{1-y}$, where $y$ is either 0 or~1.
% Therefore, the expression is simply a choice
% between $a$ for $y=1$ and $b$ for $y=0$. However, as the compiler cannot infer
% the value of $y$, both expressions $a$ and $b$ were computed and their
% resulting values were raised to the power $y$ and $1-y$ respectively.
We transformed such expressions into conditional expressions which
compute either $a$ or $b$, and avoid floating-point exponentiation.
%
% Similarly, we transformed some expressions to a more efficient form.
% For example, we replaced expressions of the form $a^{0.5}\times{b}^{0.5}$
% with $sqrt(ab)$.
Other optimizations were loop strength
reduction and common subexpression lifting.
%
% In summary, a multitude of localized optimizations were applied to achieve
% an efficient sequential baseline version, which is 1000..1500 times faster
% than the original Python code, see the next Section.
All transformations were rigorously tested to ensure they do
not modify the behavior of the algorithm.
%
\begin{comment}
Further, we replaced calls to the system's random
functions with a custom implementation of the random generator
\textit{xorshift\_128}~\cite{Marsaglia:2003:XR}. This way, random calls no longer
involve system calls, so we can support easy and fast multi-threaded random
calls by providing each thread with its private, differently seeded, random
generator.
\end{comment}

### 2017-06-01 distr-03-design.tex

This section describes how we create and evaluate a distributed version of the C++
algorithm. We didn't use the advanced intra-node parallelization techniques
described in the previous section, but found that straightforward
multi-threaded parallelization techniques sufficed: the distributed performance
is dominated by (remote) memory latency, not by node parallelism. As our
analysis will show, our pipelining optimizations will fully overlap the
parallel calculations with (pre)loading the data from remote hosts.
The details of the multi-threaded parallelism
will be discussed only where appropriate. The
distributed design follows a master-worker paradigm, where
the master controls the sequence of parallel operations and the workers perform the
calculations. For thread parallelism, we annotate the program with
OpenMP~\cite{OpenMPSpec}. For distributed communication, we use
MPI~\cite{Forum:1994:MMI:898758}.

@@

\subsubsection{Pipelining of computation and loading $\pi$}

The distributed design features two instances of pipeline parallelism. First,
at the process level, sampling of the minibatch by the master is decoupled
from \textit{update\_phi} by the workers, so the master computes the minibatch
for the next iteration while the workers compute \textit{update\_phi} for the
current iteration. Second, at the threads level within the workers, loading of
$\pi$ and \textit{update\_phi} are split up into chunks, and the calculation
in \textit{update\_phi} for the current chunk is done simultaneously with
loading $\pi$ for the next chunk. The rationale for implementing pipeline
parallelism in these places of the program, is that loading $\pi$ is the
dominant contribution within the dominant stage \textit{update\_phi}, as
will be shown in Section~\ref{eval-pipeline}.  Since loading $\pi$ is the
performance bottleneck of the distributed algorithm, performing other work
in parallel shortens the critical path.

>>>>

This section describes how we create and evaluate a distributed version of the C++
algorithm. We used the multicore CPUs in the DAS5 with straightforward
intra-node thread parallelisation, not the GPU implementations described
in the previous section. The justification is twofold: in our DAS5 system,
it is pointless to speed up the intra-node performance beyond straightforward
multithreading, because distributed performance is dominated by remote memory
latency; the analysis in Section~\ref{eval-pipeline} shows that the time for
(pre)loading the data from remote hosts greatly exceeds the calculation time.
The calculation time is fully overlapped in our pipelining implementation.
Moreover, usage of GPUs in the distributed setting would necessitate staging of
the relevant parts of $\pi/\phi$ from the host to GPU memory in each iteration,
which would eat most of the performance gain offered by the GPU because
the algorithm is so data-intensive.
% The details of the multi-threaded parallelism
% will be discussed only where appropriate.
The distributed design
follows a master-worker paradigm, where the master controls the sequence of
parallel operations and the workers perform the calculations. For thread
parallelism, we annotate the program with OpenMP~\cite{OpenMPSpec}. For
distributed communication, we use MPI~\cite{Forum:1994:MMI:898758}.

@@

\subsubsection{Pipelining of computation and loading $\pi$}
\label{sec-pipelining}

The distributed design features two instances of pipeline parallelism. First,
at the process level, sampling of the minibatch by the master is decoupled
from \textit{update\_phi} by the workers, so the master computes the minibatch
for the next iteration while the workers compute \textit{update\_phi} for the
current iteration. Second, at the threads level within the workers, loading of
$\pi$ and \textit{update\_phi} are split up into chunks, and the calculation
in \textit{update\_phi} for the current chunk is done simultaneously with
loading $\pi$ for the next chunk. The rationale for implementing pipeline
parallelism in these places of the program is that loading $\pi$ is the
dominant contribution within the dominant stage \textit{update\_phi}, as
will be shown in Section~\ref{eval-pipeline}.  Since loading $\pi$ is the
performance bottleneck of the distributed algorithm, performing other work
in parallel shortens the critical path.
